# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

```{r}
unzip("activity.zip")
activity <- read.csv("activity.csv", colClasses = c("integer", "character", "character"))
activity$time <- factor(activity$interval, levels=activity$interval[1:288])
activity$day <- strptime(activity$date, "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

Let's take a look at the daily number of steps.

```{r fig.height=4}
stepsByDate <- by(activity$steps, activity$date, sum)
hist(stepsByDate, breaks=20,
     xlab="Number of daily steps",
     main="Histogram of number of daily steps")
```

Compute mean and median:

```{r}
meanDailySteps <- mean(stepsByDate, na.rm=T); meanDailySteps
medianDailySteps <- median(stepsByDate, na.rm=T); medianDailySteps
```

## What is the average daily activity pattern?

```{r fig.height=4, fig.width=10}
stepsByInterval <- by(activity$steps, activity$time, mean, na.rm=T)
plot(stepsByInterval, type="l", xlab="Interval", ylab="Average # of steps", xaxt="n")
ticks <- seq(from=1,by=24,to=288)
axis(1,at=activity$time[ticks],labels=activity$time[ticks])
```

Which interval has the most steps on average:

```{r}
maxStepInterval <- activity$interval[which.max(stepsByInterval)]
maxStepInterval
```

## Imputing missing values

Number of missing values:

```{r}
missingValues <- is.na(activity$steps)
sum(missingValues)
```

There are `r sum(missingValues)` missing values.

`activityFilled` has missing values replaced with the average number of steps for that interval (so if a step is missing in a row and its interval is 835, then it will be filled with the average for that interval).

```{r}
missingSteps <- stepsByInterval[activity[missingValues,"time"]]
stepsFilled <- activity$steps
stepsFilled[missingValues] <- round(stepsByInterval[activity[missingValues,"time"]])
activityFilled <- cbind(activity, stepsFilled)
```

Updated Histogram:

```{r fig.height=4}
stepsByDateFilled <- by(activityFilled$stepsFilled, activity$date, sum)
hist(stepsByDateFilled, breaks=20,
     xlab="Number of daily steps",
     main="Histogram of number of daily steps with missing values filled")
```

Compute mean and median:

```{r}
meanDailyStepsFilled <- mean(stepsByDateFilled, na.rm=T)
meanDailyStepsFilled
medianDailyStepsFilled <- median(stepsByDateFilled, na.rm=T)
medianDailyStepsFilled
```

The mean is unchanged but the median is slightly less (`r medianDailyStepsFilled` is less that `r medianDailySteps`).

```{r}
totalDailySteps <- sum(stepsByDate, na.rm=T)
totalDailyStepsFilled <- sum(stepsByDateFilled, na.rm=T)
```

With imputed missing values, the total number of steps is `r totalDailyStepsFilled` compared to `r totalDailySteps` previously.


## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable for the type of day (weekday or weekend):

```{r}
activity$dayType <- factor(
  ifelse(weekdays(activity$day) %in% c("Saturday", "Sunday"),
         "weekend", "weekday")
  )
```

Plot number of steps by day type:

```{r fig.height=8, fig.width=10}
weekdayActivity <- subset(activity, activity$dayType == "weekday")
weekdayByInterval <- by(
  weekdayActivity$steps, weekdayActivity$time,
  mean, na.rm=T)

weekendActivity <- subset(activity, activity$dayType == "weekend")
weekendByInterval <- by(
  weekendActivity$steps, weekendActivity$time,
  mean, na.rm=T)

ticks <- seq(from=1,by=24,to=288)
par(mfrow=c(2,1))
par(mar=c(1,3,1,1))
plot(weekdayByInterval, type="l", xlab="", ylab="", xaxt="n", main="weekday")
par(mar=c(2,3,1,1))
plot(weekendByInterval, type="l", xlab="", ylab="", xaxt="n", main="weekend")
axis(1,at=activity$time[ticks],labels=activity$time[ticks])
```
